### 2020-08-17 text/article_vedrovice.Rmd
si + text - continue spatial analysis

The Cemetery of Vedrovice is one of the earliest and most extensively studied
cemeteries of the *Linearbandkeramik* (LBK).
Recent contributions use the evidence from the cemetery in a broader discussion
of post-marital residence patterns [@hrncir2020] and
sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
Here we focus on detailed analysis of the distribution of the grave goods and
spatial analysis of the Vedrovice cemetery in respect to gender relations.

<!-- gender relations -->

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods, especially the Bronze Age.
They argue that a the perception of gender in the Neolithic and the Bronze Age is
not the same (ibid. 142).
In response, @bickle2020 [14] shows that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

@@

<!-- site -->

<!-- sample (burials, artefacts) -->

The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities.
Thus some of the burials are excluded from the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.

Each burial was originally characterized by 55 variables but only 35 are
selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Other variables are removed because they are highly dependent on each other.

The variables fall into several categories.
Firstly there are biological and anthropological characteristics derived from
retrieved human skeletal remains including sex, age simplified into age groups,
isotope ratios or body height.

Second large category of variables consists of data on artifacts present in the
burials.
These include variables noting quantities of various types of pottery and
positions of the pottery vessels within the graves,
counts of various raw materials of chipped stone tools,
presence of polished stone axes or adzes, grinding slabs, pebbles,
bone tools and different kinds of personal ornaments.
The personal ornaments include beads made of spondylus shell or marble,
sponylus pendants or oval shaped buckles.
Presence of ochre pigment is also noted.
All the variables in this category are simplified by binarization (presence vs
absence of a trait) in order to minimize effects of only several numerous
cases on the overall pattern, for instance 20 beads in a grave <!-- 70 -->
would have much larger effect on the analysis then 1 pottery vessel in the same grave.

@@

<!-- spatial analysis -->

Data is analyzed in an R environment for statistical computing and graphics
[@rcore]  using functions of readr [@wickham2017a], dplyr [@wickham2016] and
reshape2 [@wickham2007] packages to manipulate the data.
Function daisy [@kaufman1990] from package cluster [@maechler2016] is used to
count distances between observations used in further analyses.
<!-- ADD METHODS AND PACKAGES USED IN R, or into SI?-->

<!-- software packages -->

Spatial patterning in prehistoric burial grounds is often examined to find
any internal structure, arrangement or possible grouping.
We approach the task as a point pattern analysis problem given that
a burial (event) is simplified as its centroid location in a two dimensional space.
A polygon of the excavation extent is used as a study area (a window).
Using a window closely approximating locations of the burials was considered,
e. g. buffer zone or convex/concave hull but simplified excavation polygon is
deemed the most useful because it preserves the *natural* borders and empty areas
of the original cemetery.

Overall structure of the burial ground is examined using both density based approach
and distance based methods <!--[@osullivan2010, 121-154]-->.
It is argued that density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence the location.
On the other hand the methods based on distances between neighbours and triples of neighbors
characterize the point pattern based on *second- and third-order effects*
[@nakoinz2016, 130-135; @osullivan2010, 107].
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->

In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
In case of a grave placement in a burial ground we presume that the choice of
a location is influenced predominantly by other (previous) grave locations and
is largely independent on other parameters.
In other words *first-order properties* are of little to no importance in
our point of view.
The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
are thus considered major factors in locating of the graves.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->

To characterize the pattern based on density, the quadrat test is employed.
It divides the area of interest into several cells of fixed sizes
(quadrats) and compares the count of events per quadrat with the count
expected under the null hypothesis that the underlying point process
is completely spatially random (approximated by a Poisson distribution).
Functions $G$, $F$, Ripley's $K$ [@ripley1976; @sayer2013] and $T$ [@schladitz2000] are
based on the nearest neighbour distances [@nakoinz2016, 135-144].
The $G$ function is a metric of how close events are and summarises
cummulative frequency distribution of the nearest neighbour distances.
Because of its dependence on the intensity of events in the study area, it is
subject to changes in the size of the study area.
The $F$ function, also empty-space function, is based on simulating a point pattern
and calculating of distances between random locations and the observed locations.
It is a measure of how far are observed events from any arbitrary location [@osullivan2010, 132-135].
Ripley's $K$ function is in comparison with the $G$ function resilient to
changes in the window size and works with all the proximal events up to a
certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
It is widely used in archaeology [@negre2018].
Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].

If $s_i$ is an event location, then
$G$ function is given by the formula
\@ref(eq:gfun)

\begin{equation}
  \label{eq:gfun}
  G(d) = \frac{\#[d_{min}(s_i)<d]}{n}
\end{equation}

$F$ function by
$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
Ripley's $K$ function
$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
$T$ function
$$ $$

@@

To explore whether neighboring individuals are more often of the same sex we
construct a Gabriel graph [@gabriel1969] with burial locations as nodes (Fig. \@ref(fig:gabriel)).
For each node, mean sex of the neighbors is determined and recorded.
To check how common the observed value is, we perform a permutation test
where sex of burials is randomly assigned while preserving the overall
proportion of given categories and the neighborhood struture determined by the
Gabriel graph.
The observed values of neighbors sex categories are compared against distributions
derived from the simulated data.

@@

>>>>

The Cemetery of Vedrovice is one of the earliest and most intensively studied
*Linearbandkeramik* (LBK) cemeteries.
Recent contributions use the evidence from the cemetery in a broader discussion
of post-marital residence patterns [@hrncir2020] and
sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
<!-- Here we focus on detailed analysis of the distribution of the grave goods and
spatial analysis of the Vedrovice cemetery in respect to gender relations and
social structure. -->
Here we explore the relations between burials and bodies in terms of biologicall
sex, age, localness, exceptionality of grave goods and position within the cemetery.

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods.
This is due to the notion of gender in the Neolihic that is presented as being
of a very different nature than gender in the later prehistory.
Authors show that gender in the Neolithic is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories (ibid. 140-142).
Considering the Neolithic burial evidence across Europe, authors find that it
'shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness' (ibid. 138).
In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

<!-- gender relations in LBK -->
Although gender relations in the LBK might not be

@@

<!-- site -->
The material analyzed here comes from Vedrovice – Široká u lesa cemetery.

<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities.
These burials are excluded from the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclamasclanslatorre2020.

Each burial was originally characterized by 55 variables but only subset of
35 variables was selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Other variables are removed because they are highly dependent on each other.

The variables fall into several categories.
Firstly there are biological and anthropological characteristics derived from
retrieved human skeletal remains including sex, age simplified into age groups,
isotope ratios and body height.

Second large category of variables consists of data on artifacts presence and
absence in the burials.
These include variables noting quantities of various types of pottery and
positions of the pottery vessels within the graves,
counts of various raw materials of chipped stone tools,
presence of polished stone axes or adzes, grinding slabs, pebbles,
bone tools and different kinds of personal ornaments.
The personal ornaments include beads made of spondylus shell or marble,
sponylus pendants or oval shaped buckles.
Presence of ochre pigment is also noted.
All the variables in this category are simplified by binarization (presence vs
absence of a trait) in order to minimize effects of only several numerous
cases on the overall pattern, for instance 20 beads in a grave <!-- 70 -->
would have much larger effect on the analysis then 1 pottery vessel in the same grave.

@@

<!-- spatial analysis -->
## Spatial patterning

Spatial patterning in prehistoric burial grounds is often examined to find
any internal structure, arrangement or possible grouping.
We approach the task as a point pattern analysis problem given that
a burial (event) is simplified as its centroid location in a two dimensional space.
A polygon of idealized excavation extent is used as a study area (a window).
Using a window closely approximating locations of the burials was considered,
e. g. buffer zone or convex/concave hull but simplified excavation polygon is
deemed the most useful because it preserves the borders and empty areas
of the original cemetery.

Overall structure of the burial ground is examined using both density based approach
and distance based methods. <!--[@osullivan2010, 121-154]-->
Density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence the location.
On the other hand the methods based on distances between neighbours and triples of neighbors
characterize the point pattern based on *second- and third-order effects*
[@nakoinz2016, 130-135; @osullivan2010, 107].
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->

In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
In case of a grave placement in a burial ground we presume that the choice of
a location is influenced predominantly by other (previous) grave locations and
is largely independent on other parameters.
In other words *first-order properties* are of little to no importance in
our point of view.
The *second-order properties*, and possibly *third-order properties*
are thus considered major factors in locating of the graves.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->

To characterize the pattern based on density, the quadrat test is employed.
It divides the area of interest into several cells of fixed sizes
(quadrats) and compares the count of events per quadrat with the count
expected under the null hypothesis that the underlying point process
is completely spatially random (approximated by a Poisson distribution).
Functions $G$, $F$, Ripley's $K$ [@ripley1976; @sayer2013] and $T$ [@schladitz2000] are
based on the nearest neighbour distances [@nakoinz2016, 135-144].
The $G$ function is a metric of how close events are and summarises
cummulative frequency distribution of the nearest neighbour distances.
Because of its dependence on the intensity of events in the study area, it is
subject to changes in the size of the study area.
The $F$ function, an empty-space function, is based on simulating a point pattern
and calculating of distances between random locations and the observed locations.
It is a measure of how far are observed events from any arbitrary location [@osullivan2010, 132-135].
Ripley's $K$ function is in comparison with the $G$ function resilient to
changes in the window size and works with all the proximal events up to a
certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
It is widely used in archaeology [@negre2018].
Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
For formulas of the individual functions see the supplement.

@@

To explore whether neighboring individuals are more often of the same sex we use
two methods to establish neighborhood.
The first method is based on constructing a Gabriel graph [@gabriel1969]
with burial locations as nodes (Fig. \@ref(fig:gabriel)).
Gabriel graph is a subgraph of a Delaunay triangulation

For each node, mean sex of the neighbors is determined and recorded.
To check how common the observed value is, a permutation test
where sex of burials is randomly assigned while preserving the overall
proportion of given categories and the neighborhood struture determined by the
Gabriel graph is performed.
The observed values of neighbors sex categories are compared against distributions
derived from the simulated data.

@@

The second approach of establishing neigborhood is by constructing buffer zones
around each of the burials with 6m radius.
The size of a radius is based on the results of nearest neighbors analysis, where
at the given radius there is strong evidence of a clustering process (Fig. \ref(fig:spat-funs)).
The randomness is again tested using a permutation test similar to the one
described for the neighborhood derived from the Gabriel graph.

<!-- software packages -->
Data is analyzed in an R environment for statistical computing and graphics
[@rcore]  using functions of many different packages, for the complete list
see the supplementary information.
<!-- readr [@wickham2017a], dplyr [@wickham2016] and
reshape2 [@wickham2007] packages to manipulate the data.
Function daisy [@kaufman1990] from package cluster [@maechler2016] is used to
count distances between observations used in further analyses. -->

### 2020-08-17 text/si_vedrovice.Rmd
si + text - continue spatial analysis

@@

>>>>

```{r}
library(here)
library(tidyverse)
```

# Spatial analysis

<!--
If $s_i$ is an event location, then
$G$ function is given by the formula
\@ref(eq:gfun)

\begin{equation}
\label{eq:gfun}
G(d) = \frac{\#[d_{min}(s_i)<d]}{n}
\end{equation}

$F$ function by
$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
Ripley's $K$ function
$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
$T$ function
$$ $$
-->

$s_i$ is an event location.

## G function

$$ G(d) = \frac{\sum[d_{min}(s_i)<d]}{n} $$

## F function

$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$

where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.

## Ripley's K function

$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$

@@

# Results

```{r neighborhood-sex, echo = FALSE}
knitr::include_graphics(here("plots", "nb_buffer_sex.pdf"))
knitr::include_graphics(here("plots", "nb_delaunay_sex.pdf"))
knitr::include_graphics(here("plots", "nb_gabriel_sex.pdf"))
```

```{r, echo = FALSE, message = FALSE, results = 'asis'}
b <- read_csv(here("data/temp/nb_p_buffer.csv"))
d <- read_csv(here("data/temp/nb_p_delaunay.csv"))
g <- read_csv(here("data/temp/nb_p_gabriel.csv"))

tab <- full_join(g, d, by = c("from", "to"), suffix = c(".gabriel", ".delaunay")) %>% 
  full_join(b, by = c("from", "to")) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  transmute(relation = str_c(from, to, sep = " - "),
            gabriel = mean.gabriel,
            p.gabriel = if_else(signif.gabriel, str_c(p.gabriel, "*"), as.character(p.gabriel)),
            delaunay = mean.delaunay,
            p.delaunay = if_else(signif.delaunay, str_c(p.delaunay, "*"), as.character(p.delaunay)),
            buffer = value,
            p.buffer = if_else(signif, str_c(p, "*"), as.character(p)))

knitr::kable(tab)
```
